Processing Script

### 2024-03-13 code/eda-code/eda.qmd
Cleaning
I deleted the extra exploratory-analysis file and moved some of the processing code from eda.qmd to processing-qmd, so there is only one processed datset.

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + 
  geom_histogram() + 
  ggtitle("Total Usage of No Birth Control Method per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```
The two outliers present in each of these graphs makes them very difficult to interpret. We need to explore where the outliers in these variables are coming from. China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# find the countries that represent the outliers
outliers <- mydata %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

# standardization attempt 1
mydata <- mydata %>% mutate(larcster_st = total_larcster / sum(total_larcster),
                            sarc_st = total_sarc / sum(total_sarc),
                            trad_st = total_trad / sum(total_trad),
                            none_st = total_none / sum(total_none))

# standardization attempt 2
scaling_factor_larcster <- max(mydata$total_larcster)
mydata <- mydata %>% mutate(standardized_larcster = total_larcster / scaling_factor_larcster)

# standardization attempt 3
min_count <- min(mydata$total_larcster)
max_count <- max(mydata$total_larcster)
scaled_counts <- (mydata$total_larcster - min_count) / (max_count - min_count)
print(scaled_counts)
mean_count <- mean(mydata$total_larcster)
sd_count <- sd(mydata$total_larcster)
z_score_counts <- (mydata$total_larcster - mean_count) / sd_count
```

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reprodutive age. 
```{r}
# population size standarization for total counts per birth control method
mydata <- mydata %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```

We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using LARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10)

>>>>

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + 
  geom_histogram() + 
  ggtitle("Total Usage of No Birth Control Method per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```

The outliers were explored and addressed in the processing-file.qmd in the code > processing-code folder. Now that we have standardized population variables, We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using LARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10)

### 2024-03-13 code/processing-code/processing-file.qmd
Cleaning
I deleted the extra exploratory-analysis file and moved some of the processing code from eda.qmd to processing-qmd, so there is only one processed datset.

# check if new variable was created correctly 
class(rawdata2$none_usage)
```

>>>>

# check if new variable was created correctly 
class(rawdata2$none_usage)
```
The exploratory data analysis in the eda.qmd file in the code > eda-code folder revealed outliers in the data that need to be addressed.

The two outliers make the histograms showing the distribution of total usage of birth control method by country very difficult to interpret. We need to explore where the outliers in these variables are coming from. 
```{r}
# find the countries that represent the outliers
outliers <- rawdata2 %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

# print the outliers
outliers
```

China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size.

The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# standardization attempt 1
rawdata2 <- rawdata2 %>% mutate(larcster_st = total_larcster / sum(total_larcster),
                            sarc_st = total_sarc / sum(total_sarc),
                            trad_st = total_trad / sum(total_trad),
                            none_st = total_none / sum(total_none))

# standardization attempt 2
scaling_factor_larcster <- max(rawdata2$total_larcster)
rawdata2 <- rawdata2 %>% mutate(standardized_larcster = total_larcster / scaling_factor_larcster)

# standardization attempt 3
min_count <- min(rawdata2$total_larcster)
max_count <- max(rawdata2$total_larcster)
scaled_counts <- (rawdata2$total_larcster - min_count) / (max_count - min_count)
print(scaled_counts)
mean_count <- mean(rawdata2$total_larcster)
sd_count <- sd(rawdata2$total_larcster)
z_score_counts <- (rawdata2$total_larcster - mean_count) / sd_count
```

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reproductive age. 
```{r}
# population size standardization for total counts per birth control method
rawdata2 <- rawdata2 %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```
